The response theory for the system under a non-equilibrium driving

### 2013-06-20 writedown/linear.response/paper.tex
add observation from optimal forcing

For $\eps\ge 0$ the  expectation value $O_\eps(t)$ takes the form
\begin{align}
  O_\eps(t) & =
  \int \dd\vect u f(\vect x,0)
  \int_{\mc\{\vect u,0; t\}} O[\vect X_s]\,
  \, \dd\mathcal P_\eps[\vect X_s]   \\
& = \int \dd\vect u f(\vect x,0)
  \int_{\mc\{\vect u,0; t\}} O[\vect X_s]\,
  \frac{\dd {\mathcal P}_\eps[\vect X_s]}{\dd\mathcal P_0[\vect X_s]}
  \, \dd\mathcal P_0[\vect X_s] .   
\end{align}
By using the Girsanov formula, 
\begin{align}\label{eqn:girsanov}
  \frac
  {\dd{\mathcal P_\eps}[\vect X_s]}
  {\dd\mathcal P_0[\vect X_s]}  
  =&\,
  \exp
  \big\{{\eps G[\vect X_s] - \eps^2 H[\vect X_s]}
  \big\}
\end{align}
where the short notations are defined by:
\begin{align}
  G[\vect X_s]
  &= \frac{1}\sigma\int_0^t
  \Delta F_e(s)\,\vect D(\vect q_s)\cdot\dd\vect W_s \\
  H[\vect X_s]
  &=
  \frac 1{2\sigma^2}\int_0^t
  \Delta F_e^2(s)\,\vect D^2(\vect q_s)\,\dd s.
\end{align}
Expansion of the right-hand-side of Eq.~\eqref{eqn:girsanov} in $\eps$, we have
\begin{align}\label{eqn:neq-response}
   O_\eps(t) -
  O_0(t)
  =
  \eps\, \big\langle O[\vect X_s]\cdot G[\vect X_s]\big\rangle_{0,t}
  +
  \frac{\eps^2}{2} \big\langle O[\vect X_s] \cdot (G^2[\vect X_s] - H[\vect X_s] )\big\rangle_{0,t}
  +
  \mo (\eps^3)
\end{align}
Some remarks concerning Eq.~\eqref{eqn:neq-response} seem in order:
\begin{itemize}
\item The first term on the right-hand-side of Eq.~\eqref{eqn:neq-response} is the linear
  response to the reference non-equilibrium process (driven by $F_e$, $\eps=0$). The next term is the second order response term. Obviously, one can easily write down higher orders of response by the Taylor expansion of the right-hand-side of Eq.~\eqref{eqn:girsanov}.
\item When the observable $O[\vect X_s]$ only depends on the end point of the trajectory, i.e.
  $O[\vect X_s] = O(\vect X_t)$, then $\big\langle O(\vect X_t)\cdot G[\vect X_s]\big\rangle_{0,t}$
  gives the well-know time correlation form of the linear response formula.
\item If one wants to calculate the same expectation value for a family of non-equilibrium perturbations $\Delta F_e(t)$ then 
one does not need to repeat the calculations of Eq.~\eqref{eqn:neq-response} for every member of the family. 
  If it is possible to
  express different $\Delta F_e$ in the same basis, then the responses must only be
  calculated for the single basis functions. Then with a linear combination of the responses on basis functions, one can derive
  the responses of the whole family. 
\item In the above derivation, we assume that the reference and the perturbed non-equilibrium processes
  have the same initial distribution (both of them may be different from the equilibrium distribution),
  but this is not a necessary. One can easily
  apply a similar reweighing approach to the initial distribution as we used it for the trajectory ensemble.
\end{itemize}

@@

Now we consider the optimization problem
\[
O_{min}(T) = \min_{F_e\in \mathcal{F}} O[F_e](T),
\]
where $O[F](T)$ is the expectation value of the path observable $\mathcal{O}$ that results from \eqref{eqn:langevin-1b}--\eqref{eqn:langevin-2b} with $F_e=F$ and $\eps=0$.
With the gradient being computed one can solve the optimization problem by means of gradient descent in coefficient space,
\[
a^{(i+1)}=a^{(i)} -\gamma \frac{\partial O(T)}{\partial a}.
\]

@@

\subsection{Optimal forcing}
\begin{figure}
  \centering
  \includegraphics[]{figs/fig-ctr-stat.eps}
  \caption{The optimal forcing.
    The red color indicates $\eta = 0.01$ and the
    green color indicates $\eta = 0.05$.
    In the lower plot, the dashed lines display the
    value of $P(\vect X_t)$ as the time eloves.
    The solid lines present $O[F_e](t)$.  In the upper
    plot, the solid lines present the magnitude of the
    control $F_e(t)$.
  }\label{fig:tmp7}
\end{figure}

We investigate the example of the double well potential, and
consider the problem of optimally forcing the population from the
left well to the right well. That is equal to consider the 
the following optimal forcing problem:
\begin{align}
  O_{\min}(T)& = \min_{F_e\in\mathcal F} O[F_e](T) 
\end{align}
with
\begin{align}
  O[F_e](T) &= P(\vect X_T) + \eta \int_0^T\vert F_e(t)\vect D(\vect q)\vert^2\, dt
\end{align}
where $P(\vect X_T)$ is equal to 1 when the end point of the
trajectory $\vect X_T$ is in the right well, and equal to 0 otherwise.
The integral is the ``cost'' of the control and $\eta$ is a constant
that indicates the relative magnitude of the cost. We consider the
ansatz function family $\mathcal F$ of piecewise linear functions on
$[0, T]$. 
Fig.~\ref{fig:tmp7} presents the numerical result. The magnitude of control
is shown in the upper plot, and the value of $O[F_e](T)$ and $P(\vect X_T)$
are given in the lower plot.

>>>>

For $\eps\ge 0$ the  expectation value $O_\eps(t)$ takes the form
\begin{align}
  O_\eps(t) & =
  \int \dd\vect u f(\vect x,0)
  \int_{\mc\{\vect u,0; t\}} O[\vect X_s]\,
  \, \dd\mathcal P_\eps[\vect X_s]   \\
& = \int \dd\vect u f(\vect x,0)
  \int_{\mc\{\vect u,0; t\}} O[\vect X_s]\,
  \frac{\dd {\mathcal P}_\eps[\vect X_s]}{\dd\mathcal P_0[\vect X_s]}
  \, \dd\mathcal P_0[\vect X_s] .   
\end{align}
By using the Girsanov formula, 
\begin{align}\label{eqn:girsanov}
  \frac
  {\dd{\mathcal P_\eps}[\vect X_s]}
  {\dd\mathcal P_0[\vect X_s]}  
  =&\,
  \exp
  \big\{{\eps G[\vect X_s] - \frac{\eps^2}{2} H[\vect X_s]}
  \big\}
\end{align}
where the short notations are defined by:
\begin{align}
  G[\vect X_s]
  &= \frac{1}\sigma\int_0^t
  \Delta F_e(s)\,\vect D(\vect q_s)\cdot\dd\vect W_s \\
  H[\vect X_s]
  &=
  \frac 1{\sigma^2}\int_0^t
  \Delta F_e^2(s)\,\vect D^2(\vect q_s)\,\dd s.
\end{align}
Expansion of the right-hand-side of Eq.~\eqref{eqn:girsanov} in $\eps$, we have
\begin{align}\label{eqn:neq-response}
   O_\eps(t) -
  O_0(t)
  =
  \eps\, \big\langle O[\vect X_s]\cdot G[\vect X_s]\big\rangle_{0,t}
  +
  \frac{\eps^2}{2} \big\langle O[\vect X_s] \cdot (G^2[\vect X_s] - H[\vect X_s] )\big\rangle_{0,t}
  +
  \mo (\eps^3)
\end{align}
Some remarks concerning Eq.~\eqref{eqn:neq-response} seem in order:
\begin{itemize}
\item The first term on the right-hand-side of Eq.~\eqref{eqn:neq-response} is the linear
  response to the reference non-equilibrium process (driven by $F_e$, $\eps=0$). The next term is the second order response term. Obviously, one can easily write down higher orders of response by the Taylor expansion of the right-hand-side of Eq.~\eqref{eqn:girsanov}.
\item When the observable $O[\vect X_s]$ only depends on the end point of the trajectory, i.e.
  $O[\vect X_s] = O(\vect X_t)$, then $\big\langle O(\vect X_t)\cdot G[\vect X_s]\big\rangle_{0,t}$
  gives the well-know time correlation form of the linear response formula.
\item If one wants to calculate the same expectation value for a family of non-equilibrium perturbations $\Delta F_e(t)$ then 
one does not need to repeat the calculations of Eq.~\eqref{eqn:neq-response} for every member of the family. 
  If it is possible to
  express different $\Delta F_e$ in the same basis, then the responses must only be
  calculated for the single basis functions. Then with a linear combination of the responses on basis functions, one can derive
  the responses of the whole family. 
\item In the above derivation, we assume that the reference and the perturbed non-equilibrium processes
  have the same initial distribution (both of them may be different from the equilibrium distribution),
  but this is not a necessary. One can easily
  apply a similar reweighing approach to the initial distribution as we used it for the trajectory ensemble.
\end{itemize}

@@

Now we consider the optimization problem
\[
O_{max}(T) = \max_{F_e\in \mathcal{F}} O[F_e](T),
\]
where $O[F](T)$ is the expectation value of the path observable $\mathcal{O}$ that results from \eqref{eqn:langevin-1b}--\eqref{eqn:langevin-2b} with $F_e=F$ and $\eps=0$.
With the gradient being computed one can solve the optimization problem by means of gradient descent in coefficient space,
\[
a^{(i+1)}=a^{(i)} -\gamma \frac{\partial O(T)}{\partial a}.
\]

@@

\subsection{Optimal forcing}
\begin{figure}
  \centering
  \includegraphics[]{figs/fig-ctr-stat.eps}
  \caption{The optimal forcing.  In this plot, $T=20$ ps.  The red
    color indicates $\eta = 0.01$ and the green color indicates $\eta
    = 0.05$.  In the upper plot, the solid lines present the magnitude
    of the control $F_e(t)$, which is calculated on a family of
    piecewise linear ansatz functions. The time interval of a linear
    piece is 1 ps.  In the lower plot, the dashed lines display the
    value of $P(\vect X_t)$ as the time evolves.  The solid lines
    present $O[F_e](t)$.  }\label{fig:tmp7}
\end{figure}

We investigate the example of the double well potential, and
consider the problem of optimally forcing the population from the
left well to the right well. That is equal to consider the 
the following optimal forcing problem:
\begin{align}
  O_{\max}(T)& = \max_{F_e\in\mathcal F} O[F_e](T) 
\end{align}
with
\begin{align}
  O[F_e](T) &= P(\vect X_T) - \eta \int_0^T\vert F_e(t)\vect D(\vect q)\vert^2\, dt
\end{align}
where $P(\vect X_T)$ is equal to $1$ when the end point of the
trajectory $\vect X_T$ is in the right well, and equal to 0 otherwise.
It is clear that for an unbiased double-well, the average of $P(\vect
X_T)$ is 0.5.  The integral is the ``cost'' of the control and $\eta$
is a positive constant that indicates the relative magnitude of the
cost. We consider the ansatz function family $\mathcal F$ of piecewise
linear functions on $[0, T]$.

Fig.~\ref{fig:tmp7} presents the numerical result. The magnitude of
control $F_e(t)$ is shown in the upper plot, and the value of
optimization target $O[F_e](T)$ and the value of $P(\vect X_T)$ are
given in the lower plot by the solid and dashed lines, respectively.
The plot shows that with a smaller $\eta$ (the relative magnitude of
the cost of the control is smaller), a stronger control is calculated
from the optimization, and the optimized value of $O[F_e](T)$ and
$P(\vect X_T)$ are bigger, which means the end position of $\vect X_t$
is more biased towards the right well. This observation is consistent
with intuition.  When $t<10$~ps, the magnitude of the control is
almost 0.  Near the end time $T$, the magnitude of the control firstly
quickly goes up, and then falls down a little bit.  This implies some
interesting information.  If the system were able to immediately relax
to its instantaneous equilibrium state (sometimes called
quasi-equilibrium), the population in the right well (dashed lines in
Fig.~\ref{fig:tmp7}) would immediately go down, as the control
decreases.  The fact that this does not happen, indicates that the
speed of changing the control is relatively faster that the internal
time scale of the system, so the system does not have enough time to
fully relax itself. Therefore, the observed phenomenon is truly
non-equilibrium, and our non-equilibrium linear response theory is a
tool that facilitates the investigation of this optimal forcing
problem in the non-equilibrium setting.  Intuitively, one concludes
that as the control goes up, the population in the right well actually
does not have enough time to go as high as it could be, so when the
control decreases, it continues going up rather than following the
decreasing control.
